Classical Dynamics of Harmonically Trapped Interacting Particles

### 2017-12-21 arXiv_v1/draft_10_arXiv.tex
version submitted to arXiv

@@

\usepackage{amsmath}
 \usepackage{amsfonts}
 \usepackage{amssymb}
 \usepackage{graphicx}
 \usepackage{subfigure}
% \usepackage{enumerate}

% \usepackage{verbatim}% for \begin{comment} \end{comment}

\usepackage{color}
\usepackage{xcolor}
\usepackage{bm}
\usepackage[]{hyperref}
\hypersetup{
    colorlinks = false,
    linkbordercolor = {red},
    citebordercolor = {green},
}

\usepackage{epstopdf}

@@

Motivated by the interest in the dynamics of trapped quantum gases, we study the microcanonical
dynamics of a trapped one-dimensional gas of classical particles interacting via a finite-range
repulsive force of tunable strength.  We examine two questions which have been of interest in
quantum dynamics: (1) the breathing mode (size oscillation) dynamics of the trapped gas and the
dependence of the breathing frequency on the interaction strength, and (2) the long-time relaxation
and possible thermalization of the finite isolated gas.
%
We show that the breathing mode frequency has nonmonotonic dependence on the magnitude of the mutual
repulsion, decreasing for small interactions and increasing for larger interactions.  We explain
these dependences in terms of slowing-down or speeding-up effects of two-body collision processes.
We find that the gas thermalizes within a reasonable finite timescale in the sense of
single-particle energies acquiring a Boltzmann distribution, only when the interaction strength is
large compared to the energy per particle.

@@

The non-equilibrium dynamics of isolated systems has been the subject of a large volume of work in
recent years.  In particular, in the quantum context, the study of unitary quantum evolution of
many-body and few-body systems has undergone an explosive growth.  This interest was partly fueled
by the availability of experimental systems, e.g., trapped ultracold atomic and ionic systems, where
non-dissipative dynamics can be studied explicitly due to excellent isolation from the environment.
There are also foundational reasons for studying quantum dynamics in isolation: recent activity has
led to advances in the understanding of basic principles connecting quantum dynamics, quantum chaos
and the emergence of statistical mechanics and thermalization \cite{PolkovnikovRigol_AdvPhys2016,
  BorgonoviIzrailevSantos_PhysRep2016}.  For classical many-body systems, explorations of the
connections between microscopic dynamical rules, chaos, ergodicity and statistical mechanics have a
much longer history, dating back to the time of Boltzmann and Poincare \cite{Gaspard_book_1998,
  Dorfman_book_1999, Dumas_book_KAMstory, Boltzmann_legacy_book, EckmannRuelle_RMP85,
  Gaspard_PhysicaA06}.

@@

Our first theme will be the breathing mode of the classical interacting gas, and the effect of
interactions on the frequency of the breathing mode.  In the quantum case, this has long been a
topic of interest for both bosons and fermions.  For bosons with short-range (contact) interactions,
the situation has been studied in one, two and three dimensions.  At zero interactions, the
breathing mode is twice the trapping frequency $\omega_0$ in each case, assuming isotropic trapping.
A mean field (Gross-Pitaevskii) calculation gives the frequency to be $\sqrt{D+2}$ times the
trapping frequency, where $D$ is the spatial dimension.  For three-dimensional (3D) traps, the
mean-field description is expected to be qualitatively correct, thus the breathing mode frequency is
expected to change monotonically from $2\omega_0$ at zero interactions to $\sqrt{5}\omega_0$ at
large interactions.  For 2D traps, the breathing frequency is exactly $2\omega_0$ for any
interaction strength \cite{PitaevskiiRosch_PRA97}, due to a symmetry.  Classical trapped particles
with scale-invariant interaction potentials also have breathing frequency $2\omega_0$ independent of
interaction strength \cite{PitaevskiiRosch_PRA97}.  The interaction dependence in 1D (trapped
Lieb-Liniger gas) is also remarkable: as the interaction is increased, the breathing frequency first
decreases from $2\omega_0$ to the mean-field prediction $\sqrt{3}\omega_0$, and then at larger
interactions increases again, returning to $2\omega_0$ in the infinite interaction (Tonks-Girardeau)
limit \cite{Naegerl_Science09, Tschischik_BHbreathing_PRA_2013, 1D_breathing_mode_recent,
  Bouchoule_PRL2014, KroenkeSchmelcher_BM}.
%
The breathing mode has also been investigated extensively for fermionic systems with contact
interactions \cite{GrimmSmith_unitaryfermions_PRA08, fermion2Dbreathing,
  Stringari_lowDcollective_PRA15}, and also for fermionic and bosonic systems with long-range
interactions \cite{OrignacCitro_PRA08, br_mode_long_range_interactions, Bonitz_Review}.  The
breathing mode has also been studied for a high-temperature trapped gas where a Boltzmann-equation
description is valid, both theoretically \cite{Guery-Odelin1999} and experimentally
\cite{Cornell_classsical_3Dexpt_NatPhys15}.

@@

Our second theme is the process of relaxation at longer timescales, and possible thermalization.
This topic is also motivated by intense recent research in the quantum context.  In quantum isolated
systems, thermalization is now understood in terms of the so-called eigenstate thermalization
hypothesis (ETH), which postulates that in ergodic systems the expectation values of observables in
individual eigenstates depend ony on the eigenenergy, and hence represent the thermal value
\cite{PolkovnikovRigol_AdvPhys2016, BorgonoviIzrailevSantos_PhysRep2016,
  ETH_Deutsch_Srednicki_Rigol}.  The connection to statistical mechanics is particularly difficult
for explicitly finite systems, in which case ETH is to be understood in terms of finite size scaling
\cite{Beugeling_ETHscaling_PRE14}.  In classical systems, the connection between microscopic dynamics and statistical mechanics
has been continually studied since the 19th century \cite{Dorfman_book_1999, Dumas_book_KAMstory,
  Boltzmann_legacy_book, EckmannRuelle_RMP85}.  It is generally understood that systems with nonzero
Lyapunov exponents, i.e., chaotic or ergodic systems, thermalize in the long-time limit, provided
there are enough degrees of freedom.  As in the quantum case, finite systems are particularly
intricate.  In finite systems, defining thermalization is trickier as most observables will
fluctuate or oscillate substantially.  In addition, the influence of Kolmogorov-Arnold-Moser (KAM)
orbits may play a role in preventing thermalization; it is generally believed that the fraction of
phase space where KAM physics might be relevant decreases rapidly with particle number
\cite{Dumas_book_KAMstory}.

@@

There is a significant literature on calculating Lyapunov exponents in model statistical-mechanics
systems and thus exploring ergodicity \cite{Dorfman_book_1999, deWijn_fine,
  Lyapunov_numerical_calculations, Lyapunov_analytical_calculations, FiniteTimeLyapunov}.  This
provides intuition of what types of microscopic interactions are likely to produce thermalizability,
and is an important step in the program of understanding statistical mechanics ``from the bottom
up''.  However, we are not aware of a significant literature on following explicitly the
microcanonical dynamics of statistical-mechanical or few-body systems.  Several such studies have
appeared recently \cite{DoyonSpohn_JSTAT17, Moore_arXiv17.10, JinKatsnelson_NJP13}, motivated by the
quantum dynamics literature, like the present work.  A few earlier studies have also examined
thermalization dynamics in one dimensional gravitational systems \cite{Yawn1997, Tsuchiya2000}.
%
Investigating such real-time dynamics can be expected to provide insights on the timescales and
microscopic mechanisms involved in the emergence of statistical mechanics from collections of
particles.

This paper is organized as follows: In Section \ref{sec:preparation}, we introduce the Hamiltonian,
identify and introduce the essential scales, and describe some aspects of the model and our
simulations.  The breathing frequency is treated in Section \ref{sec:breathing}.  We explain the
main features using both real-space and phase-space pictures of the interaction process, obtain
estimates for the interaction- and energy-dependence of the frequency shift, and compare these
predictions with numerics.  Section \ref{sec:relaxation} treats relaxation and thermalization.  We
propose a condition for relaxation in reasonable timescales from considerations of few-body
dynamics, compare the energy distribution with the Boltzmann distribution, show how the largest
finite-time Lyapunov exponent distribution reflects the relaxation condition, and examine shape
dynamics of the particle distribution in phase space.  Concluding remarks appear in Section
\ref{sec:concl}.

%% This paper is organized as follows: In Section \ref{section:preparation}, we perform a scaling to
%% reduce the number of parameters of the system. In addition, we briefly observe the ground state and
%% reveal two dynamically distinct regime of the system, i.e. strong interaction regime and weak
%% interaction regime. In Section \ref{section:breathing frequency}, we evaluate the breathing
%% frequency according to its mechanism, and compare that with our numerical simulation
%% result. Besides, we also introduce a rotating frame method of treating this problem, which is a
%% convenient tool to analyze simple harmonic systems. In section \ref{section:Thermalization}, we
%% focus on the thermalization property. We first examine the canonical distribution of energy to find
%% out the threshold of thermalization, i.e. under what parameter the thermalization time scale is very
%% large. In the end of this section, we measure the Lyapunov Exponent to quantify the time scales and
%% confirm the former discussion. Then we point out that there are two kind of thermalization behavior
%% that could be measured in this system, one is about energy distribution while another is about shape
%% of cloud in phase-space. Since here we find two sense of ``thermalizing", we will redefine our terms
%% -- When a system's energy distribution goes to equilibrium, we still call it ``thermalized".  But
%% when a system's distribution in phase-space goes to equilibrium, we call it ``relaxed" to distiguish
%% it from ``thermalized". So in section \ref{section:Shape} we actually study the ``thermalizing" of
%% the shape of the cloud in phase-space, or the ``relaxation" behavior of the system. There we
%% discovered a long-lasting mode, which turns out to be a quadruple mode in phase-space.

@@

We will be concerned with the size of the cloud.  We quantify the size through the root-mean-square
of particle positions $\left\lbrace x_i\right\rbrace$, 
\begin{equation}
R ~=~ \left(\overline{x^2}\right)^\frac{1}{2} ~=~ \left(\frac{1}{N}\sum_i{x_i^2}\right)^\frac{1}{2} \, ,
\label{eq:def_of_R}
\end{equation}
and refer to this quantity as the `radius' or `cloud radius'.
%
In the dilute gas limit, there are few or no interactions occuring at most times.  Thus the energy
$E$ is dominated by the trap energy, i.e., $E\approx m\omega_0^2\sum_{i}x_i^2= Nm\omega_0^2R^2$, or,
in our units,
\begin{equation}
  E\approx NR^2 \, .
  \label{eq:E_and_R}
\end{equation}
%
This approximation is good when the average distance betwen particles is much larger than the
interaction range $\sigma=1$, i.e., whenever $R\gg1$, which is the regime we consider dynamics in.
Another way of describing this dilute-gas regime is that the gas is far from the ground state
(described in the next subsection).  Neglecting the interaction energy to estimate $E$ is reasonable
even at very large interactions ($F_0\gg E/N$), because for very large interactions, the particles
behave as hard rods which bounce off each other very rapidly, so that at most instants one can
expect no collisions to be taking place, provided the gas is dilute.  It is also expected to be a
good approximation at very small $F_0$ since we can then simply neglect the interaction energy.
Thus, we will use this as a common estimate for $R$ in terms of the energy.

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[tbhp]
\centering
\includegraphics[scale=0.65]{ZhiyuPictures/N_5_GS_pre_2_rev-crop.pdf}
\caption{The ground state of a system with $N=5$ particles, shown via the positions of each
  particle, labeled 1 through 5.  At large interactions, the particles position themselves just
  outside the range of interactions of the neighboring particles.}
\label{fig:GS1}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure*}[tbph]
\includegraphics[width=0.9\textwidth]{ZhiyuPictures/stationary_and_rotating_frame_t=1-5_5T.pdf}
%
\caption{Upper (lower) panels are snapshots of the cloud configuration in phase space in the
  stationary (rotating) frame, at instants within the first trap period.  Here $N=50$, $F_0=100$,
  and $E=50000$.  }
\label{fig:Breathingfrequency2_0}
\end{figure*}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure*}[tbph]
\includegraphics[width=0.9\textwidth]{ZhiyuPictures/rotating_frame_t=0-80T.pdf}  
\caption{Phase space snapshots of the cloud dynamics viewed in the rotating frame, at longer
  timescales.  The broadening and rotation observed here are interaction effects: in the
  non-interacting gas the particles are stationary in this frame.}
\label{fig:Breathingfrequency2_1}
\end{figure*}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

Figures \ref{fig:Breathingfrequency2_0} and \ref{fig:Breathingfrequency2_1} show such phase space
snapshots.  We generally start with particles distributed at random positions around the trap
center, initially with no velocity (zero momentum).  The initial state thus has the points lined up
along the $X$ axis.  In the absence of interactions, each particle would undergo simple harmonic
oscillation with period $T=2\pi/\omega_0 = 2\pi$.  The point corresponding to each particle executes
clockwise elliptical motion around the $X$-$P$ plane.  (In our units, the elliptical trajectory is
actually circular.)  This results in the initial line remaining a straight line and rotating
clockwise with exact period $T$.  The effect of the interaction is to smear out the line and spread
the points out toward a rotationally invariant distribution in the $X$-$P$ plane.  The top panel of
Figure \ref{fig:Breathingfrequency2_0} shows this during the first period and Figure
\ref{fig:Breathingfrequency2_1} shows this process over a much longer timescale.

@@

In the absence of interactions, each point (each particle) is stationary in the rotating-frame
phase-space picture.  As seen in lower panel of Figure \ref{fig:Breathingfrequency2_0} and in Figure
\ref{fig:Breathingfrequency2_1}, interactions cause a gradual distortion of the line as well as some
degree of rotation.
%
The rotation that is visible in the already rotating $X_{\rm rot}$-$P_{\rm rot}$ frame is the
interaction-induced shift of the breathing frequency from the noninteracting value,
$\omega_B=2\omega_0$.  The distortion of the line toward an eventually rotationally invariant
distribution may be regarded as thermalization or ergodicity.  In the next sections we explore these
two interaction-induced effects.

@@

We use the Verlet algorithm (molecular dynamics) to numerically simulate the cloud, using particle
numbers between 5 and 50.  Our force is simple, so that calculating the force at each step is
inexpensive, however, the theta function dependence of the force on particle positions requires the
use of fine timesteps at the beginning and end of each collision process.

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[tbp]
\center
\includegraphics[scale=0.32]{ZhiyuPictures/freq_scanF_differentE_log_2_with_oscillation_demo.png}
\caption{Breathing mode frequency measured at different $E$ and $F_0$, for $N=5$ particles. Inset: a
  demonstration of the oscillation of $E_{trap}$, the total potential energy of particles in the
  trap, Since $E_{trap}\propto \sum_{is}{x_i^2}$, it manifests the oscillation behavior of the
  radius $R(t)$ as defined in Eq.~\ref{eq:def_of_R}.  The breathing frequency is obtained from the
  Fourier transform of $R(t)$.}
\label{fig:Breathingfrequency1}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

Without interaction, the breathing mode frequency is exactly 2.  This is visualized readily from the
top panel of \ref{fig:Breathingfrequency2_0}, where $R(t)$ is the extent of the distribution in the
horizontal ($X$) direction.  As the line rotates clockwise with frequency $\omega_0=1$, within each
period the line is twice horizontally aligned (maximum $R$) and twice vertically aligned (minimum
$R$), so that the frequency of $R(t)$ is $\omega_B=2$.  When there is interaction, the frequency
will get shifted: $\omega_B = 2+\delta$.  Numerically, we measure the radius of the cloud $R(t)$ and
get the frequency spectrum of its oscillation behavior by Fourier transform. Then we take the peak
frequency near 2 as the breathing mode frequency.

@@

The frequency measured numerically for different $E$ and $F_0$ is shown in Figure
\ref{fig:Breathingfrequency1} for a system with five particles, plotted as a function of $F_0$.  The
points each correspond to breathing-mode dynamics following from a single initial state; there is no
averaging.  The curves therefore show some noise.  However, two prominent features are clear from
these curves.  At large interactions, $F_0\gg E$, the breathing frequency is larger than $2$, and
saturates around a value which decreases with the system energy $E$.  At small interactions, the
breathing frequency is actually smaller than the non-interacting value $2$.

>>>>

\pdfoutput=1

@@

\usepackage{amsmath,amsfonts,amssymb}
\usepackage{graphicx}
\usepackage{color}

% \usepackage{enumerate}
% \usepackage{subfigure}
% \usepackage{verbatim}
% \usepackage{bm}
% \usepackage{xcolor}
% \usepackage{epstopdf}

\usepackage[breaklinks=true,colorlinks,urlcolor=blue,citecolor=blue]{hyperref}

%% \usepackage[]{hyperref}
%% \hypersetup{
%%     colorlinks = false,
%%     linkbordercolor = {red},
%%     citebordercolor = {green},
%% }

@@

Motivated by current interest in the dynamics of trapped quantum gases, we study the microcanonical
dynamics of a trapped one-dimensional gas of classical particles interacting via a finite-range
repulsive force of tunable strength.  We examine two questions which have been of interest in
quantum dynamics: (1) the breathing mode (size oscillation) dynamics of the trapped gas and the
dependence of the breathing frequency on the interaction strength, and (2) the long-time relaxation
and possible thermalization of the finite isolated gas.
%
We show that the breathing mode frequency has non-monotonic dependence on the magnitude of the
mutual repulsion, decreasing for small interactions and increasing for larger interactions.  We
explain these dependences in terms of slowing-down or speeding-up effects of two-body collision
processes.  We find that the gas thermalizes within a reasonable finite timescale in the sense of
single-particle energies acquiring a Boltzmann distribution, only when the interaction strength is
large compared to the energy per particle.

@@

The non-equilibrium dynamics of isolated systems has been the subject of a large volume of work in
recent years.  In particular, in the quantum context, the study of unitary quantum evolution of
many-body and few-body systems has undergone an explosive growth.  This interest was partly fueled
by the availability of experimental systems, e.g., trapped ultracold atomic and ionic systems, where
non-dissipative dynamics can be studied explicitly due to excellent isolation from the environment.
There are also foundational reasons for studying quantum dynamics in isolation: recent activity has
led to advances in the understanding of basic principles connecting quantum dynamics, quantum chaos
and the emergence of statistical mechanics and thermalization \cite{PolkovnikovRigol_AdvPhys2016,
  BorgonoviIzrailevSantos_PhysRep2016}.  For classical many-body systems, explorations of the
connections between microscopic dynamical rules, chaos, ergodicity and statistical mechanics have a
much longer history, dating back to the time of Boltzmann and Poincar\'e \cite{Gaspard_book_1998,
  Dorfman_book_1999, Dumas_book_KAMstory, Boltzmann_legacy_book, EckmannRuelle_RMP85,
  Gaspard_PhysicaA06}.

@@

Our first theme will be the breathing mode of the classical interacting gas, and the effect of
interactions on the frequency of the breathing mode.  In the quantum case, this has long been a
topic of interest for both bosons and fermions.  For bosons with short-range (contact) interactions,
the situation has been studied in one, two and three dimensions.  At zero interaction, the breathing
mode is twice the trapping frequency $\omega_0$ in each case, assuming isotropic trapping.  A mean
field (Gross-Pitaevskii) calculation gives the frequency to be $\sqrt{D+2}$ times the trapping
frequency, where $D$ is the spatial dimension.  For three-dimensional (3D) traps, the mean-field
description is expected to be qualitatively correct, thus the breathing mode frequency is expected
to change monotonically from $2\omega_0$ at zero interactions to $\sqrt{5}\omega_0$ at large
interactions.  Remarkably, for 2D traps, the breathing frequency is exactly $2\omega_0$ for any
interaction strength \cite{PitaevskiiRosch_PRA97}, due to a symmetry.  Classical trapped particles
with scale-invariant interaction potentials also have breathing frequency $2\omega_0$ independent of
interaction strength \cite{PitaevskiiRosch_PRA97}.  The interaction dependence in 1D (trapped
Lieb-Liniger gas) is also remarkable: as the interaction is increased, the breathing frequency first
decreases from $2\omega_0$ to the mean-field prediction $\sqrt{3}\omega_0$, and then at larger
interactions increases again, returning to $2\omega_0$ in the infinite interaction (Tonks-Girardeau)
limit \cite{Naegerl_Science09, Tschischik_BHbreathing_PRA_2013, 1D_breathing_mode_recent,
  Bouchoule_PRL2014, KroenkeSchmelcher_BM}.
%
The breathing mode has also been investigated extensively for fermionic systems with contact
interactions \cite{GrimmSmith_unitaryfermions_PRA08, fermion2Dbreathing,
  Stringari_lowDcollective_PRA15}, and also for fermionic and bosonic systems with long-range
interactions \cite{OrignacCitro_PRA08, br_mode_long_range_interactions, Bonitz_Review}.  The
breathing mode has also been studied for a high-temperature trapped gas where a Boltzmann-equation
description is valid, both theoretically \cite{Guery-Odelin1999} and experimentally
\cite{Cornell_classsical_3Dexpt_NatPhys15}.

@@

Our second theme is the process of relaxation at longer timescales, and possible thermalization.
This topic is also motivated by intense recent research in the quantum context.  In quantum isolated
systems, thermalization is now understood in terms of the so-called eigenstate thermalization
hypothesis (ETH), which postulates that in ergodic systems the expectation values of observables in
individual eigenstates depend ony on the eigenenergy, and hence represent the thermal value
\cite{PolkovnikovRigol_AdvPhys2016, BorgonoviIzrailevSantos_PhysRep2016,
  ETH_Deutsch_Srednicki_Rigol}.  The connection to statistical mechanics is particularly difficult
for explicitly finite systems, in which case ETH is to be understood in terms of finite size scaling
\cite{Beugeling_ETHscaling_PRE14}.  In classical systems, the connection between microscopic
dynamics and statistical mechanics has been continually studied since the 19th century
\cite{Gaspard_book_1998, Dorfman_book_1999, Dumas_book_KAMstory, Boltzmann_legacy_book,
  EckmannRuelle_RMP85, Gaspard_PhysicaA06}.  It is generally understood that systems with nonzero
Lyapunov exponents, i.e., chaotic or ergodic systems, thermalize in the long-time limit, provided
there are enough degrees of freedom.  As in the quantum case, finite systems are particularly
intricate.  In finite systems, defining thermalization is trickier as most observables will
fluctuate or oscillate substantially.  In addition, the influence of Kolmogorov-Arnold-Moser (KAM)
orbits may play a role in preventing thermalization; it is generally believed that the fraction of
phase space where KAM physics might be relevant decreases rapidly with particle number
\cite{Dumas_book_KAMstory}.

@@

There is a significant literature on calculating Lyapunov exponents in model statistical-mechanics
systems and thus exploring ergodicity \cite{Dorfman_book_1999, deWijn_fine,
  Lyapunov_numerical_calculations, Lyapunov_analytical_calculations, FiniteTimeLyapunov}.  This
provides intuition of what types of microscopic interactions are likely to produce thermalizability,
and is an important step in the program of understanding statistical mechanics ``from the bottom
up''.  However, we are not aware of a significant literature on following explicitly the
microcanonical dynamics of statistical-mechanical or few-body systems.  Several such studies have
appeared recently \cite{DoyonSpohn_JSTAT17, Moore_arXiv17.10, JinKatsnelson_NJP13}, motivated by the
quantum dynamics literature, like the present work.  Another line of work has examined
thermalization dynamics in one dimensional gravitational systems \cite{1Dgravitational}.
%
Investigating such real-time dynamics can be expected to provide insights on the timescales and
microscopic mechanisms involved in the emergence of statistical mechanics from collections of
particles.

This paper is organized as follows: In Section \ref{sec:preparation}, we introduce the Hamiltonian,
identify and introduce the essential scales, and describe some aspects of the model and our
simulations.  The breathing frequency is treated in Section \ref{sec:breathing}.  We explain the
main features using both real-space and phase space pictures of the interaction process, obtain
estimates for the interaction- and energy-dependence of the frequency shift, and compare these
predictions with numerics.  Section \ref{sec:relaxation} treats relaxation and thermalization.  We
propose a condition for relaxation in reasonable timescales from considerations of few-body
dynamics, compare the energy distribution with the Boltzmann distribution, show how the largest
finite-time Lyapunov exponent distribution reflects the relaxation condition, and examine shape
dynamics of the particle distribution in phase space.  Concluding remarks appear in Section
\ref{sec:concl}.

%% This paper is organized as follows: In Section \ref{section:preparation}, we perform a scaling to
%% reduce the number of parameters of the system. In addition, we briefly observe the ground state and
%% reveal two dynamically distinct regime of the system, i.e. strong interaction regime and weak
%% interaction regime. In Section \ref{section:breathing frequency}, we evaluate the breathing
%% frequency according to its mechanism, and compare that with our numerical simulation
%% result. Besides, we also introduce a rotating frame method of treating this problem, which is a
%% convenient tool to analyze simple harmonic systems. In section \ref{section:Thermalization}, we
%% focus on the thermalization property. We first examine the canonical distribution of energy to find
%% out the threshold of thermalization, i.e. under what parameter the thermalization time scale is very
%% large. In the end of this section, we measure the Lyapunov Exponent to quantify the time scales and
%% confirm the former discussion. Then we point out that there are two kind of thermalization behavior
%% that could be measured in this system, one is about energy distribution while another is about shape
%% of cloud in phase space. Since here we find two sense of ``thermalizing", we will redefine our terms
%% -- When a system's energy distribution goes to equilibrium, we still call it ``thermalized".  But
%% when a system's distribution in phase space goes to equilibrium, we call it ``relaxed" to distiguish
%% it from ``thermalized". So in section \ref{section:Shape} we actually study the ``thermalizing" of
%% the shape of the cloud in phase space, or the ``relaxation" behavior of the system. There we
%% discovered a long-lasting mode, which turns out to be a quadruple mode in phase space.

@@

We will be concerned with the size of the cloud.  We quantify the size through the root-mean-square
of particle positions $\left\lbrace x_i\right\rbrace$, 
\begin{equation}
R ~=~ \left(\overline{x^2}\right)^\frac{1}{2} ~=~ \left(\frac{1}{N}\sum_i{x_i^2}\right)^\frac{1}{2} \, ,
\label{eq:def_of_R}
\end{equation}
and refer to this quantity as the `radius' or `cloud radius'.
%
In the dilute gas limit, there are few or no interactions occuring at most times.  Thus the energy
$E$ is dominated by the trap energy, i.e., $E\approx m\omega_0^2\sum_{i}x_i^2= Nm\omega_0^2R^2$, or,
in our units,
\begin{equation}
  E ~\approx~ NR^2 \, .
  \label{eq:E_and_R}
\end{equation}
%
This approximation is good when the average distance betwen particles is much larger than the
interaction range $\sigma=1$, i.e., whenever $R\gg1$, which is the regime we consider dynamics in.
Another way of describing this dilute-gas regime is that the gas is far from the ground state
(described in the next subsection).  Neglecting the interaction energy to estimate $E$ is reasonable
even at very large interactions ($F_0\gg E/N$), because for very large interactions, the particles
behave as hard rods which bounce off each other very rapidly, so that at most instants one can
expect no collisions to be taking place, provided the gas is dilute.  It is also expected to be a
good approximation at very small $F_0$ since we can then simply neglect the interaction energy.
Thus, we will use this as a common estimate for $R$ in terms of the energy.

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[tbhp]
\centering
\includegraphics[scale=0.65]{N_5_GS_pre_2_rev-crop.pdf}
\caption{The ground state of a system with $N=5$ particles, shown via the positions of each
  particle, labeled 1 through 5.  At large interactions, the particles position themselves just
  outside the range of interactions of the neighboring particles.}
\label{fig:GS1}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure*}[tbph]
\includegraphics[width=0.9\textwidth]{stationary_and_rotating_frame_t_1-5_5T.pdf}
%
\caption{Upper (lower) panels are snapshots of the cloud configuration in phase space in the
  stationary (rotating) frame, at instants within the first trap period.  Here $N=50$, $F_0=100$,
  and $E=50000$.  }
\label{fig:Breathingfrequency2_0}
\end{figure*}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure*}[tbph]
\includegraphics[width=0.9\textwidth]{rotating_frame_t_0-80T.pdf}  
\caption{Phase space snapshots of the cloud dynamics viewed in the rotating frame, at longer
  timescales.  The broadening and rotation observed here are interaction effects: in the
  non-interacting gas the particles are stationary in this frame.}
\label{fig:Breathingfrequency2_1}
\end{figure*}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

Figures \ref{fig:Breathingfrequency2_0} and \ref{fig:Breathingfrequency2_1} show such phase space
snapshots.  We generally start with particles distributed at random positions around the trap
center, initially with no velocity (zero momentum).  The initial state thus has the points lined up
along the $X$ axis.  In the absence of interactions, each particle would undergo simple harmonic
oscillation with period $T=2\pi/\omega_0 = 2\pi$.  The point corresponding to each particle executes
clockwise elliptical motion around the $X$-$P$ plane.  (In our units, the elliptical trajectory is
actually circular.)  This results in the distribution retaining its linear shape and rotating
clockwise with exact period $T$.  The effect of the interaction is to smear out the line and spread
the points out toward a rotationally invariant distribution in the $X$-$P$ plane.  The top panel of
Figure \ref{fig:Breathingfrequency2_0} shows this during the first period and Figure
\ref{fig:Breathingfrequency2_1} shows this process over a much longer timescale.

@@

In the absence of interactions, each point (each particle) is stationary in the rotating-frame phase
space picture.  As seen in the lower row of Figure \ref{fig:Breathingfrequency2_0} and in Figure
\ref{fig:Breathingfrequency2_1}, interactions cause a gradual distortion of the line as well as some
degree of rotation.
%
The rotation that is visible in the already rotating $X_{\rm rot}$-$P_{\rm rot}$ frame is the
interaction-induced shift of the breathing frequency $\omega_B$ from the noninteracting value,
$2\omega_0$.  The distortion of the line toward an eventually rotationally invariant distribution
may be regarded as thermalization or ergodicity.  In the next sections we explore these two
interaction-induced effects.

@@

We use the Verlet algorithm (molecular dynamics) to numerically simulate the cloud, using particle
numbers between 5 and 50.  Our force is simple, so that calculating the force at each step is
inexpensive, however, the theta function dependence of the force on particle positions requires the
use of fine timesteps at the beginning and end of each collision process.  The simulation is purely
microcanonical: No external bath or thermalizing mechanisms are introduced.

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[tbp]
\center
\includegraphics[scale=0.32]{freq_scanF_differentE_log_2_with_oscillation_demo.png}
\caption{Breathing mode frequency measured at different $E$ and $F_0$, for $N=5$ particles. Inset: a
  demonstration of the oscillation of $E_{trap}$, the total potential energy of particles in the
  trap, Since $E_{trap}\propto \sum_{is}{x_i^2}$, it manifests the oscillation behavior of the
  radius $R(t)$ as defined in Eq.~\ref{eq:def_of_R}.  The breathing frequency is obtained from the
  Fourier transform of $R(t)$.}
\label{fig:Breathingfrequency1}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

Without interaction, the breathing mode frequency $\omega_B$ is exactly 2.  This is visualized
readily from the top panel of Figure \ref{fig:Breathingfrequency2_0}, where $R(t)$ is the extent of
the distribution in the horizontal ($X$) direction.  As the line rotates clockwise with frequency
$\omega_0=1$, within each period the line is twice horizontally aligned (maximum $R$) and twice
vertically aligned (minimum $R$), so that the frequency of $R(t)$ is $\omega_B=2$.  When there is
interaction, the frequency will get shifted: $\omega_B = 2+\delta$.  Numerically, we measure the
radius of the cloud $R(t)$ and get the frequency spectrum of its oscillation behavior by Fourier
transform. Then we take the peak frequency near 2 as the breathing mode frequency.

@@

The frequency measured numerically for different $E$ and $F_0$ is shown in Figure
\ref{fig:Breathingfrequency1} for a system with five particles, plotted as a function of $F_0$.  The
points each correspond to breathing-mode dynamics following from a single initial state; there is no
averaging.  The curves therefore show some noise.  However, two prominent features are clear from
these curves.  At large interactions, $F_0\gg E$, the breathing frequency is larger than $2$, and
saturates around a value which decreases with the system energy $E$.  At small interactions, the
breathing frequency is smaller than the non-interacting value $2$.

Below, we will provide a detailed phase space argument for these behaviors.  However, a simple
real-space picture explains both effects qualitatively as well.  For very large $F_0$, when the
interaction is `hard-core'-like, two particles exchange momentum instantaneously during a collision.
By exchanging the labels of the particles during the collision, this can be interpreted as follows:
particle $A$ carrying momentum $P_A$ jumps by distance $\sigma$ to the right, while particle $B$
carries its momentum $P_B$ and jumps by distance $\sigma$ to the left.  In this manner, every
collision will save a particle some time, $\frac{\sigma}{v}$.  This translates into an increase of
the breathing frequency.  The speed $v$ per particle is on average $\sim\sqrt{E/N}$, and the number
of collision each particle experiences in one period of harmonic oscillation is $\sim N$; hence we
obtain $\delta\sim N^{\frac{3}{2}}E^{-\frac{1}{2}}$.  When $F_0$ is smaller, we can no longer think
of the particles as hard rods; the collisions now take finite time during which the speeds of the
particles are slowed down and then sped up again, as they either cross paths or bounce from each
other.  When $F_0$ is small enough, this approaching time will at some critical value of $F_0$
consume the time saved by the finite range of the interaction.  This explains why there is a
small-$F_0$ regime for which the shift $\delta$ is negative.  Figure \ref{fig:Breathingfrequency1}
shows that this critical value of $F_0$, where $\delta$ changes sign and $\omega_B$ crosses $2$,
increases with energy.
